---
title: "Predictive Modeling - DO mg/L trends"
format: 
  html:
    toc: true
    code-fold: true
warning: false
message: false
error: true
embed-resources: true

---

# Setup  

```{r}
library(tidyverse)
library(MuMIn)
library(doParallel)
library(glmmTMB)
library(tictoc)
library(vegan)
library(ggfortify)
library(performance)
```

```{r}
source(here::here("R", "Analyses_for_paper",
                  "05_predictive_modeling",
                  "050_setup.R"))
```


Natural-log transforming nutrient medians and turbidity, to ensure linearity of effects and stabilize variance.  

Centering and standardizing (1SD) predictors to make sure model converges and aid in relative interpretation (comparing effect sizes among variables on very different scales); once we have a top model set/averaged model, we can back-transform and interpret slopes in their raw units.  


```{r}
# pick the specific predictors we've agreed on
# center and scale all but response

dat_domgl <- dat_all |> 
  select(reserve,
         domgl_trend,
         # lat/temp/par PCA
         tpld_PC1,
         # wq medians
         spcond_median, turb_median.log, 
         # wq trends
         temp_trend, spcond_trend, turb_trend,
         # nut medians
         chla_median.log, nh4_median.log, no23_median.log, po4_median.log,
         # nut trends
         chla_trend, nh4f_trend, no23f_trend, po4f_trend,
         # met
         precp_median, precp_trend, dailyPAR_trend) |> 
  mutate(across(tpld_PC1:last_col(),
                function(x) as.vector(scale(x))))

formula_domgl <- paste0("domgl_trend ~ ", paste(names(dat_domgl[3:ncol(dat_domgl)]), collapse = " + "), " + (1|reserve)")

# generate formula without random effect
# due to singular fit when including random effect
formula_fixed_domgl <- paste0("domgl_trend ~ ", paste(names(dat_domgl[3:ncol(dat_domgl)]), collapse = " + "))
```

# Models     

## Create and evaluate global models    

### DO mgl

```{r}
mod_domgl <- glmmTMB(as.formula(formula_domgl),
                    data = dat_domgl)
performance::check_singularity(mod_domgl)
summary(mod_domgl)$varcor

# drop random factor due to singular fit and extremely low variance in random effect
mod_domgl <- lm(as.formula(formula_fixed_domgl),
                data = dat_domgl)
performance::check_singularity(mod_domgl)

r2(mod_domgl)
check_collinearity(mod_domgl)
summary(mod_domgl)
```



#### Evaluate diagnostics - original model  

```{r}
# assign model at top so this code chunk can be reused
mod = mod_domgl


# this code shows a variety of plots, which is nice
performance::check_model(mod)

# what are those VIFs
sort(car::vif(mod), decreasing = TRUE)
# temp, do, no23 medians.


# we'll look at other plots ourselves

# pull out pearson residuals
resids <- resid(mod, type = "pearson")

op <- par()

# pull out predictor data frame
preds_df <- mod$model[2:ncol(mod$model)]

# set up a 4-plot layout
par(mfrow = c(2, 2))

# plot the main diagnostic graphs
plot(resids ~ fitted.values(mod),
     xlab = "fitted",
     ylab = "residual")
abline(h = 0, col = "red3")
hist(resids, breaks = 20)
qqnorm(resids)
qqline(resids, col = "gray60", lty = 2)

# set up a bigger plot layout
par(mfrow = c(4, 5))

# plot residuals against every predictor
for(i in 1:ncol(preds_df)){
  plot(resids ~ preds_df[[i]],
       xlab = names(preds_df)[i],
       ylab = "pearson residuals")
  abline(h = 0, col = "red3")
}

# go back to normal plot layout
par(op)

```

