---
title: "Cluster analsyis of NERRS SWMP water quality national dataset (station medians)"
output: html_document
author: For Questions, Dave Parrish (parrishd@vims.edu)
date: "2025-05-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(htmltools)
library(tidyverse)
library(psych)
library(dplyr)
library(cluster)
library(factoextra)
library(ggrepel)
library(RColorBrewer)
```

## NERR Cluster

```{r}
nerr_h_comp<-read.csv('data/WQ-NUT_overallMedians.csv')
head(nerr_h_comp)
colnames(nerr_h_comp)


nerr_h_comp$L10_turb <- log10(nerr_h_comp$turb_median)
nerr_h_comp$L10_po4f <- log10(nerr_h_comp$po4f)
nerr_h_comp$L10_no23f <- log10(nerr_h_comp$no23f)
nerr_h_comp$L10_nh4f <- log10(nerr_h_comp$nh4f)
nerr_h_comp$L10_chla <- log10(nerr_h_comp$chla)
nerr_h_comp <- nerr_h_comp %>% 
         dplyr::select(station,spcond_median, temp_median, do_mgl_median, ph_median, L10_turb, L10_chla, L10_nh4f, L10_no23f,L10_po4f)
nerr_h_comp <- nerr_h_comp %>% 
       rename(code = 'station',Temp='temp_median',SpCond='spcond_median',Turb='L10_turb',DO='do_mgl_median',NO23='L10_no23f',
              NH4='L10_nh4f',PO4='L10_po4f',Chla='L10_chla',pH=ph_median)

colnames(nerr_h_comp)
pca_nerr = prcomp(nerr_h_comp[,2:ncol(nerr_h_comp)], center = TRUE, scale = TRUE)
pca_nerr
pca_nerr$rotation
pca_nerr$x

## rotate the pca (arbitrary, but wanted to conform visualizations with previous work)
pca_nerr$rotation[, c(2,3)] <- pca_nerr$rotation[, c(2,3)] * -1
pca_nerr$x[, c(2,3)] <- pca_nerr$x[, c(2,3)] * -1
pca_nerr$
pca_nerr
pca_nerr$rotation
write.csv(pca_nerr$rotation, 'dataout/swmp_pca_loadings_spc.csv')
pca_nerr$x

#pca_nerr <- prcomp(df.n, center=FALSE, scale=FALSE)
pca_smry<-summary(pca_nerr)
pca_smry
# str(pca_smry)
pca_smry$importance
nerr_col<-'#05829C'
fviz_eig(pca_nerr,barfill = nerr_col,xlab = 'Principle Components',
  barcolor = nerr_col,main='',)+   
            theme(axis.text=element_text(size=16),
            axis.title=element_text(size=18,face="bold"))

```


## Kaiser Criterion helps to establish number of Principle Components that explain enough variance. According to the criterion, one should use as many PCs, as there is number of eigenvalues which are higher than 1. (3 PCs in this case)

```{r}
eig.val<-get_eigenvalue(pca_nerr)
eig.val
```




```{r}
for(i in 1:3){
  var<-get_pca_var(pca_nerr)
  print(var$contrib)
  plot(fviz_contrib(pca_nerr, "var", axes=i, xtickslab.rt=90,fill = nerr_col,
                    color = nerr_col)+   
            labs(title=paste("Contributions of variables to PC",i)) +
            theme(
              title=element_text(size=18),
              axis.text=element_text(size=16),
              axis.title=element_text(size=18,face="bold")), main = 'PC1')
}
```



```{r out.extra='style="max-width:none; width:100vw; margin-left:calc(50% - 50vw);"', fig.width = 10, fig.height =10, device = "svg"}
v<-c(1:3)
v.pair<-combn(v, 2)
v.pair
class(summary(pca_nerr))
str(summary(pca_nerr))
smry.pca_nerr<-summary(pca_nerr)
prop.var<-smry.pca_nerr$importance[2,]

for(c in 1:ncol(v.pair)){      
  pair<-v.pair[,c]
  xlbl<-paste0('PC',pair[1],' (',round(prop.var[pair[1]]*100,digits=1),'%)')
  ylbl<-paste0('PC',pair[2],' (',round(prop.var[pair[2]]*100,digits=1),'%)')
  pl<-fviz_pca_var(pca_nerr,axes=pair,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,     # Avoid text overlapping
             title = " ",
             labelsize = 10
             )+ coord_fixed() + xlab(xlbl) +  
             ylab(ylbl) +
             theme(axis.text=element_text(size=28),
             axis.title=element_text(size=32,face="bold"))
  print(pl)
}

```

```{r out.extra='style="max-width:none; width:100vw; margin-left:calc(50% - 50vw);"', fig.width = 10, fig.height =10, device = "svg"}
v<-c(1:3)
v.pair<-combn(v, 2)
v.pair

for(c in 1:ncol(v.pair)){
  pair<-v.pair[,c]
  xlbl<-paste0('PC',pair[1],' (',round(prop.var[pair[1]]*100,digits=1),'%)')
  ylbl<-paste0('PC',pair[2],' (',round(prop.var[pair[2]]*100,digits=1),'%)')
  pl<-fviz_pca_var(pca_nerr,axes=pair,
             #col.var = "contrib", # Color by contributions to the PC
             #radient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,     # Avoid text overlapping
             title = " ",
             labelsize = 10
             )+ coord_fixed() +  xlab(xlbl) +  
             ylab(ylbl) + 
            theme(axis.text=element_text(size=28),
            axis.title=element_text(size=32,face="bold"))
  print(pl)
}

```
```{r out.extra='style="max-width:none; width:100vw; margin-left:calc(50% - 50vw);"', fig.width = 10, fig.height =10, device = "svg"}
stn_ids<-read.csv('data/stn_map_ids.csv',stringsAsFactors = F)
nerr_h_comp.stn_ids.mrg<-merge(nerr_h_comp,stn_ids,by.x='code',by.y='code',all.x=T)

fviz_pca_biplot(pca_nerr, geom='point',
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )+
  geom_text(aes(label=nerr_h_comp.stn_ids.mrg$Map.Code),hjust=1.2, vjust = 1.2,col='gray60')+ coord_fixed()

nerr_h_comp.stn_ids.pc.mrg<-cbind(nerr_h_comp.stn_ids.mrg,pca_nerr$x)
write.csv(nerr_h_comp.stn_ids.pc.mrg,'dataout/stn_med_pc_spc.csv')
```


```{r}
pca_nerr_transform = as.data.frame(pca_nerr$x[,1:3])
```


How many clusters?

```{r}
k.max <- 15 # we want to look at the within squares for clusters 2-15. 
wss <- sapply(1:k.max, function(k){kmeans(pca_nerr_transform, k, nstart = 50)$tot.withinss})
plot(1:k.max, wss, type = "b", pch = 19, frame = FALSE, #type = b indicates points joined by lines
     xlab = "Number of CLusters K",  #pch is the shape of the point, 19 = circle 
     ylab = "Total Within Clusters Sum of Squares")
```


```{r}
fviz_nbclust(pca_nerr_transform, kmeans, method = 'wss')
```



```{r}
fviz_nbclust(pca_nerr_transform, kmeans, method = 'silhouette')
#fviz_nbclust(pca_nerr_transform, kmeans, method = 'gap_stat')
```


## Although unsupervised metrics such as silhouette width suggested setting the K number of clusters for this dataset to three, four clusters were selected to capture a fourth, generally eutrophic cluster characterized by higher phosphate than the other eutrophic cluster that was generally characterized by high nitrate + nitrite.

```{r out.extra='style="max-width:none; width:100vw; margin-left:calc(50% - 50vw);"', fig.width = 10, fig.height =10, device = "svg"}

v<-c(1:3)
pc.pair<-combn(v, 2)

n_k<-4 # set 4 clusters

for (k in n_k){ #change to range if want to look at multiple clusters
  set.seed(42)
  kmeans_nerr = kmeans(pca_nerr_transform, centers = k, nstart = 50)
  pc.labs<-colnames(kmeans_nerr$centers)
  clstr.lab<-LETTERS[1:k]
  clusters<-1:k
  #clust_clr<-brewer.pal(n = k, name = "Set2")
  
  kmeans_nerr$cluster[kmeans_nerr$cluster == 1]<-6
  kmeans_nerr$cluster[kmeans_nerr$cluster == 2]<-5
  kmeans_nerr$cluster[kmeans_nerr$cluster == 3]<-8
  kmeans_nerr$cluster[kmeans_nerr$cluster == 4]<-7
  
    
  kmeans_nerr$cluster[kmeans_nerr$cluster == 5]<-1 
  kmeans_nerr$cluster[kmeans_nerr$cluster == 6]<-3 
  
  
  kmeans_nerr$cluster[kmeans_nerr$cluster == 7]<-2 
  kmeans_nerr$cluster[kmeans_nerr$cluster == 8]<-4 
  
  
  clust_clr<-c("#66C2A5", "#FC8D62", "#8DA0CB",'#6E260E')
  clust_clr
  for(c in 1:ncol(pc.pair)){
    pair<-v.pair[,c]
    xlbl<-paste0('PC',pair[1],' (',round(prop.var[pair[1]]*100,digits=1),'%)')
    ylbl<-paste0('PC',pair[2],' (',round(prop.var[pair[2]]*100,digits=1),'%)')
    pc.lab<-pc.labs[pair]
    p<-fviz_cluster(kmeans_nerr, data = pca_nerr_transform[,pair], geom='point',stand=F, choose.vars=pc.lab,
      ggtheme = theme_minimal(),  axes = pair) +
      geom_label_repel(size =5,aes(label = nerr_h_comp.stn_ids.mrg$Map.Number, colour=cluster),
                       max.overlaps = Inf,show.legend = F) + coord_fixed() +
      scale_colour_manual(values = clust_clr,labels=clstr.lab) +
      scale_fill_manual(values = clust_clr,labels=clstr.lab) +
      scale_shape_manual(values=c(16,17,15,18),labels=clstr.lab) +
      xlab(xlbl) +
      ylab(ylbl) +
      theme(axis.text=element_text(size=28),
      axis.title=element_text(size=32,face="bold"),
      legend.title=element_text(size=22), 
      legend.text=element_text(size=20))
    
    print(p)
    
  
  
  }
  #
  stn_groups<-data.frame(nerr_h_comp$code,kmeans_nerr$cluster) 
  
  
  
  #join clusters with station metadata
  colnames(stn_groups)<-c('code','cluster')
  
  stn_groups2<-merge(stn_groups,nerr_h_comp,by='code')
  
  head(stn_groups2)
  unique(stn_groups2$cluster)
  
  stn_groups2$cluster <- factor(stn_groups2$cluster, levels = clusters)
  levels(stn_groups2$cluster)<-clstr.lab
  for(prm in colnames(nerr_h_comp)[2:ncol(nerr_h_comp)]){
    # Basic box plot
    p <- ggplot(stn_groups2, aes_string(x='cluster', y=prm, fill='cluster')) + 
      geom_boxplot(alpha=.6) + scale_fill_manual(values=clust_clr)+ scale_x_discrete("cluster", drop = FALSE) +
      theme(axis.text=element_text(size=32),
            axis.title=element_text(size=34,face="bold"),
            legend.title=element_text(size=24), 
            legend.text=element_text(size=22))
    print(p)
    
  }
  data_for_map<-merge(stn_groups2,nerr_h_comp.stn_ids.mrg, on=station)
  write.csv(data_for_map,paste0('dataout/data_for_map_spc',k,'clusters_spc.csv'),row.names=F)
  data_for_swmp_synth_group<-merge(stn_groups2[,1:2],nerr_h_comp.stn_ids.pc.mrg, on=code)
  head(data_for_swmp_synth_group)
  data_for_swmp_synth_group
  
  data_for_swmp_synth_group  <-data_for_swmp_synth_group %>% dplyr::select("code","cluster","SpCond","Temp",
                                            "DO","pH","Turb","Chla","NH4","NO23","PO4",
                                            "PC1","PC2","PC3","PC4","PC5","PC6","PC7","PC8","PC9")
  write.csv(data_for_swmp_synth_group,'dataout/swmp_clstr_med_pc_stations_spc.csv')
}

```
