---
title: "PCA"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE, 
                      error = TRUE,
                      fig.width = 7,
                      fig.height = 7)
```

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(vegan)
library(ggplot2)
```

# Data import and shaping  

```{r}
load(here::here("Outputs", "calculated_trends", "long-term-trends.RData"))
trends_all <- bind_rows(trends_df, trends_nut)  # not met trends, because stations don't match wq/nut

# can probably join in primary SWMP station based on reserve code, for later iterations
```

```{r}
all_wide <- trends_all |> 
  mutate(station = substr(station, start = 1, stop = 5),
         sig_trend = case_when(sig_trend == "yes" ~ 1,
                               sig_trend == "no" ~ 0,
                               is.na(sig_trend) ~ NA_real_),
         sig_seasonality = case_when(sig_seasonality == "yes" ~ 1,
                                     sig_seasonality == "no" ~ 0,
                                     is.na(sig_seasonality) ~ NA_real_)) |> 
  select(station, parameter, Slope_overall = Slope,
         Spring_estimate, Fall_estimate,  # possible in ordination or explanatory
         Winter_estimate, Summer_estimate,
         overall_median, sig_trend, sig_seasonality) |>   # explanatory  
  pivot_wider(id_cols = station,
              names_from = parameter,
              values_from = Slope_overall:sig_seasonality,
              names_glue = "{parameter}_{.value}",
              names_vary = "slowest") |> 
  tibble::column_to_rownames("station")
```

```{r}
# select only chl and DO params of interest, and only their slopes
# and widen out the data frame
response_vars <- c("chla_n", "do_mgl_median",
                        "do_proportion_below2", "do_proportion_below5")

# extract
slopes_only <- all_wide |> 
  select(all_of(paste0(response_vars, "_Slope_overall"))) 
# scale
slopes_scaled <- as.data.frame(apply(slopes_only, MARGIN = 2, scale))
rownames(slopes_scaled) <- rownames(slopes_only)

# replace NAs with 0s

# we lose 30 stations to that na.omit - I suspect a lot of them are because DO proportion
# trends couldn't be calculated - that was problematic
# yep 24 rows with NA in proportion below 2 slope; actually 23 for chl

# going to substitute 0 for when a slope couldn't be calculated
# but to avoid skewing the scaling, will only add into scaled data frame
# RUN THIS BY OTHERS.......
slopes_scaled[][is.na(slopes_scaled[])] <- 0
```

```{r}
# pull out a few additional factors for exploration
# seasonal slopes, temp trend (air and water),
# overall medians for all params
env_factors <- all_wide |> 
  select(ends_with("overall_median"),
         temp_median_Slope_overall)
names(env_factors) <- stringr::str_remove(names(env_factors), "median_")
# and scale
env_scaled <- as.data.frame(apply(env_factors, MARGIN = 2, scale))
rownames(env_scaled) <- rownames(env_factors)

```


# PCA  

Using the `vegan` package because you can do fancier things with it after you've run the PCA. I scaled all columns manually during data processing because the output from `vegan`'s `scale` argument seems to give ever so slightly different results than `scale` in the base R functions `princomp` and `prcomp`. By scaling manually, I am ensuring the math comes out the same.  

```{r}
pca_out <- rda(slopes_scaled)
```

## Output  

```{r}
eigs <- summary(eigenvals(pca_out))
knitr::kable(eigs,
             caption = "Variance explained by PC axes",
             digits = 4)
```


```{r}
knitr::kable(scores(pca_out, 
                    choices = 1:4,
                    display = "species",
                    scaling = 0),
             caption = "unscaled loadings onto PC axes",
             digits = 4)
```


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"))
```

We fed 4 variables into this PCA, and can see that the first two axes explain about 83% of the variation in the data.  

What we see from both the loadings and the biplot is that our DO trends separate sites along Axis 1, and the chla trends separate sites along Axis 2.  

Unsurprisingly, increasing "time hypoxic" (as represented by do_proportion_below_2 or 5 mg/L) is almost directly opposite increasing median DO mg/L.  

I do not see any groupings of stations in this output.  

## Compare NMDS  

```{r}
mds_out <- metaMDS(slopes_scaled, distance = "euclidean",
                   autotransform = FALSE)

# get species scores using envfit
# see https://stackoverflow.com/questions/57393901/how-to-get-species-score-for-ordination-with-metamds
spp_scores <- envfit(mds_out, slopes_scaled)
```

```{r}
plot(mds_out, type = "text")
plot(spp_scores, add = TRUE, cex = 0.8, col = "red")
```

```{r}
# try to make it the same orientation as the PCA
mds_scores <- scores(mds_out, tidy = TRUE)
spp_scores2 <- as.data.frame(scores(spp_scores, display = "vectors"))

mds_scores_rev <- mds_scores |> 
  mutate(across(c("NMDS1", "NMDS2"),
                function(x) x*-1))

spp_scores_rev <- spp_scores2 |> 
  mutate(across(c("NMDS1", "NMDS2"),
                function(x) x*-1)) |> 
  rownames_to_column("var")

ggplot() +
  geom_text(data = mds_scores_rev, 
            aes(x = NMDS1, y = NMDS2, label = label),
            size = 3) +
  geom_segment(data = spp_scores_rev,
               aes(x = 0, y = 0, 
                   xend = NMDS1*4, yend = NMDS2*4),
               arrow = grid::arrow(),
               col = "red3") +
  geom_text(data = spp_scores_rev, 
            aes(x = NMDS1*4, y = NMDS2*4, 
                label = var),
            col = "red3") +
  ggtitle("NMDS") +
  theme_bw()
```


```{r}
# now do the PCA one in ggplot
# try to make it the same orientation as the PCA
pca_scores <- scores(pca_out, tidy = TRUE) |> 
  filter(score == "sites")
pca_spp_scores <- scores(pca_out, tidy = TRUE) |> 
  filter(score == "species")


ggplot() +
  geom_text(data = pca_scores, 
            aes(x = PC1, y = PC2, label = label),
            size = 3) +
  geom_segment(data = pca_spp_scores,
               aes(x = 0, y = 0, 
                   xend = PC1*0.8, yend = PC2*0.8),
               arrow = grid::arrow(),
               col = "red3") +
  geom_text(data = pca_spp_scores, 
            aes(x = PC1, y = PC2, 
                label = label),
            col = "red3") +
  ggtitle("PCA") +
  theme_bw()
```



## PCA - More detail  

The following graph is the same as above, but instead of points, each station is labelled with its station code (minus the `wq` or `nut` suffix).  


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "text"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"))
```


Now we can pick a few sites and see if we're interpreting correctly.  

-  `gndbl` looks like it should have a relatively high increase in chla and not much trend in DO; maybe a small increase.  
-  `welht` is at the other end - potentially large decrease in chla, and potential decrease in DO (with an increasing number of hypoxic events).  
-  `cbmmc` seems to have little trend in chla, and increasing DO/decreasing hypoxic events.  
-  `apaeb` looks fairly in the middle for chl trends, and on the decreasing DO side.  

```{r, fig.width = 5, fig.height = 5}
check_dat <- trends_all |> 
  filter(parameter %in% c("chla_n", "do_mgl_median",
                          "do_proportion_below2",
                          "do_proportion_below5"))
check_stns <- check_dat |>
  mutate(station = substr(station, start = 1, stop = 5)) |> 
  filter(station %in% c("gndbl", "welht", "cbmmc", "apaeb"))

ggplot() +
  geom_histogram(data = check_dat, aes(x = Slope),
                 fill = "gray",
                 col = "gray20") +
  geom_vline(xintercept = 0, col = "black", linewidth = 1) +
  geom_vline(data = check_stns, 
             aes(col = station, xintercept = Slope),
             linewidth = 1) +
  facet_wrap(~parameter, scales = "free") +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") +
  labs(title = "histograms of long-term trends, with specific stations overlaid")
```

The thoughts about each station generally check out. 




# Adding additional overlays  

The following were generated using `vegan::envfit()`. These variables have also already been centered and scaled. I'm approaching this using a single type of metric at a time because there are so many parameters.   


## Seasonally calculated slopes of our main responses  

```{r}
# select vars
env_factors1 <- all_wide |> 
  select(starts_with(response_vars) & ends_with("estimate"))

# clean up names
names(env_factors1) <- stringr::str_remove(names(env_factors1), "_estimate")
names(env_factors1) <- stringr::str_remove(names(env_factors1), "_proportion")
names(env_factors1) <- stringr::str_remove(names(env_factors1), "_median")

# and scale
env1_scaled <- as.data.frame(apply(env_factors1, MARGIN = 2, scale))
rownames(env1_scaled) <- rownames(env_factors1)

# substitute 0 in place of NAs  
env1_scaled[][is.na(env1_scaled[])] <- 0


# env fit
ev1 <- envfit(pca_out, env1_scaled)
```


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "Seasonal Trends in responses overlaid")
plot(ev1, cex = 0.6)
```


Seeing a bit of separation from hypoxic events in fall and winter. These all have significant correlations with the PC axes; chl ones are more correlated with PC1, and DO ones with PC2. We can see in the PC2 numbers for do_below_etc the ones that break away diagonally from the axes in the triplot.    

```{r}
ev1
```


## Seasonally calculated slopes of predictors  

```{r}
# select vars
env_seas_preds <- all_wide |> 
  select(ends_with("_estimate") &
         starts_with(c("temp", "spcond", "nh4f", "po4f")),
         ends_with("_Slope_overall") &
         starts_with(c("temp", "spcond", "nh4f", "po4f")))

# clean up names
names(env_seas_preds) <- stringr::str_remove(names(env_seas_preds), "_estimate")
names(env_seas_preds) <- stringr::str_remove(names(env_seas_preds), "_median")
names(env_seas_preds) <- stringr::str_remove(names(env_seas_preds), "_Slope")
# and scale
envSeas_scaled <- as.data.frame(apply(env_seas_preds, MARGIN = 2, scale))
rownames(envSeas_scaled) <- rownames(env_seas_preds)

# substitute 0 in place of NAs  
envSeas_scaled[][is.na(envSeas_scaled[])] <- 0


# env fit
evSeasPreds <- envfit(pca_out, envSeas_scaled)
```

```{r}
tempSeas <- envfit(pca_out, select(envSeas_scaled, starts_with("temp")))

biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "Temperature slopes overlaid")
plot(tempSeas, cex = 0.6)
```

Winter temperature slopes seem to separate out some sites along PC1.  

```{r}
spcondSeas <- envfit(pca_out, select(envSeas_scaled, starts_with("spcond")))

biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "SpCond slopes overlaid")
plot(spcondSeas, cex = 0.6)
```

Winter and Fall slopes, a little bit.  

```{r}
nh4Seas <- envfit(pca_out, select(envSeas_scaled, starts_with("nh4")))

biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "NH4 slopes overlaid")
plot(nh4Seas, cex = 0.6)
```

Winter slopes going in a pretty different direction from the others, along both axes. I don't remember if I accounted for censoring when calculating seasonal trends - **need to check this**.  


```{r}
po4Seas <- envfit(pca_out, select(envSeas_scaled, starts_with("po4")))

biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "PO4 slopes overlaid")
plot(po4Seas, cex = 0.6)
```

PO4 separating out mostly along axis 2 (chl). Again it's Winter that seems most different from others.  

# Reminder  

Remember, all we're doing here is looking at how our sites separate out along these DO and chl gradients. This doesn't necessarily indicate drivers, but does have some interesting implications for changes in temp, nh4, and po4 in Winter, and their associations with DO and chla change.  

```{r}
evSeasPreds
```



## Latitude  

Seems like just coloring by latitude is a simpler way to look at its effect for now. I'll use ggplot in a rough way for this.  

```{r}
pca_species <- as.data.frame(scores(pca_out, display = "species"))
pca_sites <- as.data.frame(scores(pca_out, display = "sites")) |> 
  rownames_to_column(var = "station")

latitudes <- readr::read_csv(here::here("helper_files", "sampling_stations.csv")) |> 
  select(station = "Station Code", "Latitude") |> 
  mutate(station = substr(station, start = 1, stop = 5)) |> 
  filter(station %in% rownames(all_wide))

pca_sites <- left_join(pca_sites, latitudes)
```

```{r, fig.width = 5, fig.height = 5}
ggplot(pca_sites, aes(x = PC1, y = PC2, col = Latitude)) +
  geom_point(size = 3, alpha = 0.6) +
  # scale_color_distiller(palette = "PuBuGn") +
  scale_color_viridis_c() +
  geom_segment(data = pca_species,
               aes(x = 0, y = 0,
                   xend = PC1, yend = PC2),
               arrow = grid::arrow(),
               col = "red3") +
  labs(title = "Latitude as color")
```


## Seasonality amplitudes  

```{r}
load(here::here("Outputs", "calculated_trends", "seasonal_characterization.RData"))
```


```{r}
seasonal_amps <- seasonal_by_stn |> 
  mutate(ampl = case_when(parameter == "chla_n" ~ median_norm_ampl,
                          .default = median_ampl),
         station = substr(station, start = 1, stop = 5)) |> 
  select(station, parameter, ampl) |> 
  filter(station %in% rownames(slopes_scaled)) |> 
  pivot_wider(names_from = parameter,
              names_prefix = "ampl_",
              values_from = ampl) |> 
  column_to_rownames("station")

names(seasonal_amps) <- stringr::str_remove(names(seasonal_amps), "_median")
names(seasonal_amps) <- stringr::str_remove(names(seasonal_amps), "_proportion")

# and scale
env_amps <- as.data.frame(apply(seasonal_amps, MARGIN = 2, scale))
rownames(env_amps) <- rownames(seasonal_amps)

# substitute 0 in place of NAs  
env_amps[][is.na(env_amps[])] <- 0
```

```{r}
# env fit
ev_amps <- envfit(pca_out, env_amps)
```


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "Seasonal amplitudes of key parameters overlaid")
plot(ev_amps, cex = 0.6)
```

```{r}
ev_amps
```


## Overall parameter medians  

```{r}
# select vars
env_factors2 <- all_wide |> 
  select(ends_with("overall_median"))

# clean up names
names(env_factors2) <- stringr::str_remove(names(env_factors2), "_overall_median")
names(env_factors2) <- stringr::str_remove(names(env_factors2), "_proportion")
names(env_factors2) <- stringr::str_remove(names(env_factors2), "_median")

# and scale
env2_scaled <- as.data.frame(apply(env_factors2, MARGIN = 2, scale))
rownames(env2_scaled) <- rownames(env_factors2)

# substitute 0 in place of NAs  
env2_scaled[][is.na(env2_scaled[])] <- 0


# env fit
ev2 <- envfit(pca_out, env2_scaled)
```


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "Overall station medians of all parameters overlaid")
plot(ev2, cex = 0.6)
```


Some overall medians have strong correlations with our axes, and others don't. I'm not entirely sure how to interpret these outputs.  

```{r}
ev2
```



## Long-term slopes of other parameters  

```{r}
# select vars
env_factors3 <- all_wide |> 
  select(ends_with("Slope_overall"),
         -starts_with(response_vars))

# clean up names
names(env_factors3) <- stringr::str_remove(names(env_factors3), "_Slope_overall")
names(env_factors3) <- stringr::str_remove(names(env_factors3), "_median")

# and scale
env3_scaled <- as.data.frame(apply(env_factors3, MARGIN = 2, scale))
rownames(env3_scaled) <- rownames(env_factors3)

# substitute 0 in place of NAs  
env3_scaled[][is.na(env3_scaled[])] <- 0


# env fit
ev3 <- envfit(pca_out, env3_scaled)
```


```{r}
biplot(pca_out,
       display = c("sites",
                   "species"),
       type = c("text",
                "point"),
       xlab = paste0("PC1 (", round(eigs[2, 1]*100, 2), "%)"),
       ylab = paste0("PC2 (", round(eigs[2, 2]*100, 2), "%)"),
       main = "Long-term slopes of other parameters overlaid")
plot(ev3, cex = 0.6)
```

Temperature is kind of key here, based on our conversations. It correlates more with PC2 (chla trends), but not strongly (r2 = 0.07, see below). Only DO % has a decent r2, which isn't surprising. Also unsurprising that it's most correlated with PC1, the DO axis.  

```{r}
ev3
```