---
title: ""
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      error = TRUE,
                      fig.height = 5,
                      fig.width = 7)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggiraph)
source(here::here("QAQC_flags", "functions.R"))
```

```{r}
# define which flags to keep: currently just 0, 4, -4, 5
flags_keep <- c("<0>", "<4>", "<-4>", "<5>")

# path to data; change this when files get moved to a better location
path <- here::here("Data_processing", "compiled_by_stn")
```

# About this html file  

-  Headings are WQ/NUT, and the parameters we're using are each a single tab under which type of data they are.  
-  Code is present, but folded. If you want to unfold a portion of code, click on the `code` button on the right.  
-  This file was made by using 2 rmarkdown files - one as a 'parent', and one called a 'child' that is called as many times as I need it to be - i.e., for each parameter. So the code chunks will mostly look the same because it's just the same ones being called over and over again.    



# Days with data  

-  I'm only working with 1 station as an example, but in the end the goal is to have multiple stations showing for the same parameter in the same tab.  
-  Hover over bars on the plot to see the Year/Month and # days of {useable/bad} data points.  
    -  a "useable" day was defined as a day where at least one data point was flagged okay, below detection, historical, or corrected.  
    -  a "bad" day was a day where there were 0 data points in the above flags.  
    -  I find the "bad" days easier to spot problems in, which is why I put it first.... but open to input.  

# WQ {.tabset} 

```{r}
# go through each of the WQ+NUT stations
# that meet certain criteria:
# currently active, with start date in or before 2011
# both wq and nut at same station
# for now pulling data from Data_processing subfolder but 
# eventually it will live in the main 'data' folder

mstns <- readr::read_csv("sampling_stations.csv") %>%
  janitor::clean_names() %>%   #read.csv threw an error for some reason
  tidyr::separate(active_dates, into = c("start_date", "first_ end_date"),
                  sep = "-",
                  fill = "right",
                  extra = "drop") %>%  # only want the first date if there are multiples
  filter(lubridate::decimal_date(lubridate::my(start_date)) <= 2011.0,
         status == "Active")

# match up wq/nut intersection for stations that meet time length criteria
stns_wq <- str_sub(grep("wq$", mstns$station_code, value = TRUE),
                   end = -3)
stns_nut <- str_sub(grep("nut$", mstns$station_code, value = TRUE),
                   end = -4)


stns_wq_nut <- intersect(stns_wq, stns_nut)
  
rm(mstns, stns_nut, stns_wq)
```


Loop through each station to produce graphs (and stats if needed)  

```{r}
# check .RData files, only loop over stations that we have data for
fls <- dir(path, pattern = ".RData$")

toUse <- stns_wq_nut[paste0(stns_wq_nut, "wq.RData") %in% fls & paste0(stns_wq_nut, "nut.RData") %in% fls]


for(st in toUse){
  cat(st)
  # this sort of works, but the named data frame is also being loaded into the environment, which I don't want
  assign("wq", get(load(paste0(path, "/", st, "wq.RData"))))
  assign("nut", get(load(paste0(path, "/", st, "nut.RData"))))
}
```



## Temp  

```{r}
parm <- "temp"
```

```{r, child = "counting_data_child.Rmd"}

```


## SpCond  

```{r}
parm <- "spcond"
```

```{r, child = "counting_data_child.Rmd"}

```


## Sal  

```{r}
parm <- "sal"
```

```{r, child = "counting_data_child.Rmd"}

```

## DO pct  

```{r}
parm <- "do_pct"
```

```{r, child = "counting_data_child.Rmd"}

```


## DO mgl  


```{r}
parm <- "do_mgl"
```

```{r, child = "counting_data_child.Rmd"}

```




# NUT {.tabset}  

I screwed this up because I treated it the same way as WQ data. Only grab samples are included; not diel. So we never expect lots of data points in a given day....  

```{r}
rm(dat)
dat <- import_local(here::here("data"), "gndblnut",
                    collMethd = 1)
st <- "gndblnut"
```


## Chl *a*  


```{r}
parm <- "chla_n"
```

```{r, child = "counting_data_child.Rmd"}

```

## PO4  

```{r}
parm <- "po4f"
```

```{r, child = "counting_data_child.Rmd"}

```

## NH4  

```{r}
parm <- "nh4f"
```

```{r, child = "counting_data_child.Rmd"}

```

## NO23  

```{r}
parm <- "no23f"
```

```{r, child = "counting_data_child.Rmd"}

```

## NO3  

excluding for now because GND doesn't determine this separately  

## NO2  

excluding for now because GND doesn't determine this separately  


