---
title: "Combining predictors V3"
output: 
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

re-run 5/29/24 after using natural-log transformation on nutrients for trends (rather than log10)  

re-run again 7/8/24, to insert 0s for DO<2 trend at stations where that trend did not calculate due to all values being 0  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 8.5,
                      fig.width = 8)

# packages
library(tidyverse)


# parameters and columns to keep
params_wq <- c("temp_median", "spcond_median", 
               "turb_median", "do_mgl_median",
               "do_pct_median")
params_nut <- c("chla_n", "chla_n_cens",
                "nh4f", "nh4f_cens",
                "no23f", "no23f_cens",
                "po4f", "po4f_cens")
params_met <- c("totprcp_total",
                "dailyPAR_median")
params_trends <- c("temp_median", "spcond_median", "turb_median",
                   "do_mgl_median", "do_pct_median", "do_proportion_below2",
                   "chla_n", "po4f_mdl", "nh4f_mdl", "no23f_mdl",
                   "dailyPAR_median",
                   "sqrt_precp_total")

# stations we'll remove due to being too deep or problematic in some other way
# but where trends were mostly calculated, so we'll keep them until the end

#ADD LKS Stations
remove_stns <- c("pdbgd", "kachd", "kacsd", "sfbfm","lksbl", "lksol", "lksba", "lkspo")
```

Need to bring everything together: slopes and long-term medians.  

# Data import  

Read in the previously calculated things. Only keep key parameters (and medians). Find primary SWMP MET station for each reserve and bind to the wq/nut stations on reserve.  

```{r load-data}
# medians
# wq/nut
medians_wqnut <- read.csv(here::here("Outputs", 
                                     "01_calculated_medians",
                                     "WQ-NUT_overallMedians.csv"))

# met
medians_met <- read.csv(here::here("Outputs", 
                                   "01_calculated_medians",
                                   "MET_overallMedians.csv"))

# seasonal amplitude
#medians_seasonality <- read.csv(here::here("Outputs", 
 #                                          "03_calculated_seasonality",
  #                                         "seasonal_amplitude_medians.csv"))


# long-term trends
# wq/nut/met
trends_wqnutmet <- read.csv(here::here("Outputs", 
                                               "02_calculated_long-term-trends", "bam_outputs_MDL",
                                              "long-term-trends.csv"))

# seasonal amplitude
#trends_seasonality <- read.csv(here::here("Outputs", 
 #                                         "03_calculated_seasonality",
  #                                        "seasonal_amplitude_trends.csv"))


# station info
mdat <- readr::read_csv(here::here("helper_files", "sampling_stations.csv"))

```

# Filtering and selecting  

Determine primary SWMP MET station at each reserve. Some reserves have more than one; select the one with the longest time series.  

```{r}
primary_met <- mdat |> 
  filter(isSWMP == "P",
         str_ends(`Station Code`, "met")) |> 
  select(met_stn = `Station Code`) |> 
  mutate(reserve = substr(met_stn, 1, 3),
         station = substr(met_stn, 1, 5))  
  
# use reserve column to join these to other stns
# station to match with trends table

# reserves_with_2 <- janitor::get_dupes(primary_met, reserve)
# 
# trends_met |> 
#   filter(station %in% reserves_with_2$met_stn) |> 
#   select(station, ts_start, ts_end, ts_length, n_points) |> 
#   distinct() |> 
#   arrange(station, ts_start) |> 
#   View()

# only one met station per reserve in the trends file
# so just choose based on the trends file

primary_met <- primary_met |> 
  filter(met_stn %in% unique(trends_wqnutmet$station))

# make sure all primary met stations appear in the met medians file
# we want the statement below to be true
# sum(!(primary_met$station %in% medians_met$station)) == 0
```

```{r}
# trends
trnds <- trends_wqnutmet |> 
  select(-X) |> 
  mutate(reserve = substr(station, start = 1, stop = 3),
         station = substr(station, start = 1, stop = 5)) |> 
  filter(parameter %in% params_trends) |> 
  mutate(parameter = str_remove(parameter, "_median"),
         parameter = case_when(parameter == "sqrt_precp_total" ~ "precp",
                               parameter == "chla_n" ~ "chla",
                               .default = parameter),
         parameter = paste0(parameter, "_trend")) |> 
  select(reserve, station, parameter, Slope)

intcpt <- trends_wqnutmet |> 
  select(-X) |> 
  mutate(reserve = substr(station, start = 1, stop = 3),
         station = substr(station, start = 1, stop = 5)) |> 
  filter(parameter %in% params_trends) |> 
  mutate(parameter = str_remove(parameter, "_median"),
         parameter = case_when(parameter == "sqrt_precp_total" ~ "precp",
                               parameter == "chla_n" ~ "chla",
                               .default = parameter),
         parameter = paste0(parameter, "_intcpt")) |> 
  select(reserve, station, parameter, Intercept)

# pivot wider, to one row per station
trnds_water <- trnds |> 
  filter(parameter %in% c("temp_trend", "spcond_trend", "turb_trend",
                          "do_mgl_trend", "do_pct_trend", "do_proportion_below2_trend",
                          "chla_trend", "po4f_mdl_trend", "nh4f_mdl_trend",
                          "no23f_mdl_trend")) |> 
  pivot_wider(id_cols = c(reserve, station),
              names_from = parameter,
              values_from = Slope) |> 
  rename(domgl_trend = do_mgl_trend)

intrcpt_water <- intcpt |> 
  filter(parameter %in% c("temp_intcpt", "spcond_intcpt", "turb_intcpt",
                          "do_mgl_intcpt", "do_pct_intcpt", "do_proportion_below2_intcpt",
                          "chla_intcpt", "po4f_mdl_intcpt", "nh4f_mdl_intcpt",
                          "no23f_mdl_intcpt")) |> 
  pivot_wider(id_cols = c(reserve, station),
              names_from = parameter,
              values_from = Intercept) |> 
  rename(domgl_intcpt = do_mgl_intcpt)

trend_intrcpt<- 
      left_join(trnds_water, intrcpt_water, by = c("reserve","station"))

trnds_met <- trnds |> 
  filter(parameter %in% c("dailyPAR_trend", "precp_trend")) |> 
  pivot_wider(id_cols = c(reserve, station),
              names_from = parameter,
              values_from = Slope) 

intcpt_met <- intcpt |> 
  filter(parameter %in% c("dailyPAR_intcpt", "precp_intcpt")) |> 
  pivot_wider(id_cols = c(reserve, station),
              names_from = parameter,
              values_from = Intercept) 

trend_intrcpt_met<- 
      left_join(trnds_met, intcpt_met, by = c("reserve","station"))


# trim the water medians file
medians_wqnut <- medians_wqnut |> 
  select(station, Latitude,
         temp_median,
         spcond_median,
         domgl_median = do_mgl_median,
         dopct_median = do_pct_median, 
         turb_median,
         chla_median = chla,
         nh4f_median = nh4f,
         no23f_median = no23f,
         po4f_median = po4f)

# trim the met medians file
medians_met <- medians_met |> 
  filter(station %in% primary_met$station) |> 
  mutate(reserve = substr(station, 1, 3)) |> 
  select(reserve, 
         precp_median = totprcp_total,
         dailyPAR_median)
```


```{r}
write.csv(trend_intrcpt, here::here("Outputs", 
                                     "04_compiled_predictors",
                                     "trend_MDL.csv"),
          row.names = FALSE,
          na = "")


write.csv(trend_intrcpt_met, here::here("Outputs", 
                                     "04_compiled_predictors",
                                     "trend_MDL_met.csv"),
          row.names = FALSE,
          na = "")
```

# Combining  - NOT UPDATED

```{r}
# these joins will cut out stations where trends weren't calculated
water_all <- left_join(trnds_water, medians_wqnut,
                       by = "station")

met_all <- left_join(trnds_met, medians_met,
                     by = "reserve") |> 
  left_join(primary_met, by =c("station","reserve")) |> 
  select(-station)

# get wq, nut, met all together again
all_predictors <- left_join(water_all, met_all,
                            by = "reserve") |> 
  relocate(met_stn, .after = station) |> 
  relocate(Latitude, .after = met_stn) |> 
  rename(latitude = Latitude)


# weed out the problem stations - too deep, or missing trend(s)  
all_predictors <- all_predictors |> 
  filter(!(station %in% remove_stns))


# in the stations that are left, any missing do below 2 trends
# are because the time series was all 0s - insert 0 for trend in these
all_predictors <- all_predictors |> 
  mutate(do_proportion_below2_trend = case_when(is.na(do_proportion_below2_trend) ~ 0,
                                                .default = do_proportion_below2_trend))
```

# Writing out  

```{r}
write.csv(all_predictors, here::here("Outputs", 
                                     "04_compiled_predictors",
                                     "compiled_predictors_MDL_slopes.csv"),
          row.names = FALSE,
          na = "")
```

