---
title: "Summary of long-term trends"
subtitle: "Using trends of log-transformed nutrients - not log10"
date: today
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    embed-resources: true
echo: true
warning: false
message: false
error: true
fig-height: 7
fig-width: 8
---

As of 5/29/2024, calculating trends on natural-log transformed nutrients, so that (per Gelman et al. 2021, "Regression and Other Stories") coefficients can be interpreted as "approximate proportional differences; with a coefficient of 0.05, a difference of 1 in x corresponds to an approximate 5% difference in y, and so forth."  

# Setup  

```{r}
library(tidyverse)
```


Read in the output from `02_long-term-trend__calculation_analyses.Rmd` - this file has all the standard errors, p-values, and everything else that didn't get incorporated into the `compiled predictors` data frame.  

```{r}
trnds <- read.csv(here::here("Outputs", "02_calculated_long-term-trends",
                    "long-term-trends.csv"))
```

Remove certain stations; those below are removed during data frame combining.    

```{r}
# stations we'll remove due to being too deep or problematic in some other way
# but where trends were mostly calculated, so we'll keep them until the end
remove_stns <- c("pdbgd", "kachd", "kacsd", "sfbfm")

# need to affix wq and nut to them though  
remove_stns <- c(paste0(remove_stns, "wq"), paste0(remove_stns, "nut"))

trnds <- trnds |> 
  filter(!(station %in% remove_stns))
```



```{r}
#| eval: false


# See if any reserves have more than one met station
# they don't, so don't need to evaluate this chunk
# keeping it to document that we checked
trnds |> 
  select(station) |> 
  filter(str_ends(station, "met")) |> 
  distinct() |> 
  janitor::tabyl(station)
```


# Histograms and numerical summary  

```{r}
ggplot(trnds,
       aes(x = Slope,
           fill = sig_trend)) +
  geom_histogram(col = "gray40",
                 alpha = 0.8,
                 bins = 30) +
  facet_wrap(~parameter,
             scales = "free") +
  geom_vline(xintercept = 0,
             linewidth = 1) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Histograms of slope (change/yr) for each parameter",
       fill = "p < 0.05?")
```


```{r}
trnds_summ <- trnds |> 
  filter(!is.na(Slope)) |> 
  summarize(.by = parameter,
            n = n(),
            median_slope = median(Slope),
            iqr_slope = IQR(Slope),
            mean_slope = mean(Slope),
            sd_slope = sd(Slope),
            quantiles = list(round(quantile(Slope), 4))) |> 
  mutate(across(median_slope:sd_slope,
                function(x) round(x, 4)))
```

```{r}
DT::datatable(trnds_summ)
```


# Ordered slope plots  

```{r}
p <- trnds |>
  arrange(parameter, Slope, station) |> 
  mutate(stparm = paste0(station, parameter)) |> 
  ggplot(aes(y = forcats::fct_inorder(stparm), x = Slope)) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high,
                      col = sig_trend),
                  alpha = 0.8) +
  geom_vline(xintercept = 0) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(title = paste("slopes for all stations"),
       subtitle = "sorted from lowest to highest",
       x = "slope estimate and 95% ci",
       y = "",
       col = "p < 0.05?") +
  theme(legend.position = "bottom")

p +
  theme(axis.text.y = element_blank())

p +
  coord_flip() +
  theme(axis.text.x = element_blank())
```

