---
title: "Seasonal amplitudes and trends"
output: 
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      error = TRUE,
                      fig.width = 7,
                      fig.height = 5)
```

This Rmd file is living in the `R/Long-Term_trend_analysis` directory because we are looking for changes through time in seasonal amplitude. Amplitudes were originally calculated in `R/Short-Term_seasonality/seasonality_viz` and are re-calculated here for completeness (and not having to search for output from that folder now that we are also interested in a long-term component).  

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

source(here::here("helper_files", "definitions.R"))
source(here::here("helper_files", "functions.R"))

load(here::here("Data", "QAQCd_monthly_byType", "SWMP_monthlyWQ.RData"))
load(here::here("Data", "QAQCd_monthly_byType", "SWMP_monthlyNUT.RData"))
# load(here::here("Data", "QAQCd_monthly_byType", "SWMP_monthlyMET.RData"))

wq <- wq |> 
  mutate(do_proportion_below2 = round(doLessThan2_total / doLessThan2_nValid, 4),
         do_proportion_below5 = round(doLessThan5_total / doLessThan5_nValid, 4)) 
```

```{r}
# we really only want to use stations where we've already calculated long-term trends
# so read those in too and subset the main data frames
load(here::here("Outputs", "calculated_trends", "long-term-trends.RData"))

# station names so far are the long form (wq and nut at ends)
wq <- filter(wq, station %in% trends_df$station)
nut <- filter(nut, station %in% trends_nut$station)
```



```{r}
# annualize is a custom function
chl_annual <- annualize(nut, chla_n) |> 
  select(station, year, an_normalized_amplitude) |> 
  nest(data = c(year, an_normalized_amplitude))

do_annual <- annualize(wq, do_mgl_median) |> 
  select(station, year, an_normalized_amplitude) |> 
  nest(data = c(year, an_normalized_amplitude))

tmp_annual <- annualize(wq, temp_median) |> 
  select(station, year, an_normalized_amplitude) |> 
  nest(data = c(year, an_normalized_amplitude))
```

## Model  

Currently this is just a simple linear model; does not account for autocorrelation in residuals  

```{r}
mod_fun <- function(df){
  lm(an_normalized_amplitude ~ year, data = df)
}
```


### Chla  

```{r, fig.width = 5, fig.height = 7}
par(mfrow = c(2, 1))
chl_mods <- chl_annual |> 
  mutate(mod = purrr::map(data, mod_fun),
         tidy = purrr::map(mod, broom::tidy))

chl_estimates <- chl_mods |> 
  select(station, tidy) |> 
  unnest(cols = tidy) |> 
  filter(term == "year") |> 
  select(-term)

hist(chl_estimates$estimate, breaks = 50)
abline(v = 0, col = "blue", lwd = 2)
hist(chl_estimates$p.value, breaks = 50)
abline(v = 0.05, col = "red", lwd = 2)
```

### DO mg/L  

```{r, fig.width = 5, fig.height = 7}
par(mfrow = c(2, 1))
do_mods <- do_annual |> 
  mutate(mod = purrr::map(data, mod_fun),
         tidy = purrr::map(mod, broom::tidy))

do_estimates <- do_mods |> 
  select(station, tidy) |> 
  unnest(cols = tidy) |> 
  filter(term == "year") |> 
  select(-term)

hist(do_estimates$estimate, breaks = 50)
abline(v = 0, col = "blue", lwd = 2)
hist(do_estimates$p.value, breaks = 50)
abline(v = 0.05, col = "red", lwd = 2)
```

### water temp  

Change in normalized seasonal amplitude of water temperature

```{r, fig.width = 5, fig.height = 7}
par(mfrow = c(2, 1))
tmp_mods <- tmp_annual |> 
  mutate(mod = purrr::map(data, mod_fun),
         tidy = purrr::map(mod, broom::tidy))

tmp_estimates <- tmp_mods |> 
  select(station, tidy) |> 
  unnest(cols = tidy) |> 
  filter(term == "year") |> 
  select(-term)

hist(tmp_estimates$estimate, breaks = 50)
abline(v = 0, col = "blue", lwd = 2)
hist(tmp_estimates$p.value, breaks = 50)
abline(v = 0.05, col = "red", lwd = 2)
```

# prepare these for joining with other predictors  

```{r}
tmp_estimates$param = "temp"
do_estimates$param = "do_mgl"
chl_estimates$param = "chla"

all_estimates <- bind_rows(chl_estimates, do_estimates, tmp_estimates) |> 
  mutate(test = "trend in normalized seasonal amplitude")

saveRDS(all_estimates, file = here::here("Outputs", "calculated_trends",
                                         "seasonal_amplitude_trends.rds"))
```

# Diagnostics  

I'm not crazy about how residuals look from a linear model, which generated the information above. I started down this road to check for autocorrelation in the residuals, which doesn't seem like a major problem, but it's worth looking at the data and what seems to be going on.  

### Chl  

```{r, results = FALSE, fig.height = 7, fig.width = 7}
op <- par()
par(mfrow = c(2, 2))

chl2 <- chl_annual |> unnest(cols = data)
for(i in seq_along(unique(chl2$station))){
  st = unique(chl2$station)[i]
  tmp <- chl2 |> filter(station == st)
  print(plot(an_normalized_amplitude ~ year, data = tmp, main = paste(st, "time series"), type = "b"))
  lm1 <- lm(an_normalized_amplitude ~ year, data = tmp)
  print(plot(lm1, which = 1, main = st))
  print(plot(lm1, which = 2, main = st))
  print(acf(resid(lm1), main = st))
  cat("\n-----------------------")
  cat("\n-----------------------\n")
}

# not seeing systemic autocorrelation issues, actually
# a few stations might have some
# but since we're not overly focused on -pvalues, I'm not
# sure it's worth the effort to incorporate

# i wasn't crazy about residual plots though
```

### Chl - log transformed??  

```{r, results = FALSE, fig.height = 7, fig.width = 7}
op <- par()
par(mfrow = c(2, 2))

chl2 <- chl_annual |> unnest(cols = data)
for(i in seq_along(unique(chl2$station))){
  st = unique(chl2$station)[i]
  tmp <- chl2 |> filter(station == st)
  print(plot(log10(an_normalized_amplitude) ~ year, data = tmp, main = paste(st, "time series"), type = "b"))
  lm1 <- lm(log10(an_normalized_amplitude) ~ year, data = tmp)
  print(plot(lm1, which = 1, main = st))
  print(plot(lm1, which = 2, main = st))
  print(acf(resid(lm1), main = st))
  cat("-----------------------")
  cat("-----------------------")
}

# not seeing systemic autocorrelation issues, actually
# a few stations might have some
# but since we're not overly focused on -pvalues, I'm not
# sure it's worth the effort to incorporate

# i wasn't crazy about residual plots though
```


### DO mg/L  

```{r, results = FALSE, fig.height = 7, fig.width = 7}
op <- par()
par(mfrow = c(2, 2))

do2 <- do_annual |> unnest(cols = data)
for(i in seq_along(unique(do2$station))){
  st = unique(do2$station)[i]
  tmp <- do2 |> filter(station == st)
  print(plot(an_normalized_amplitude ~ year, data = tmp, main = paste(st, "time series"), type = "b"))
  lm1 <- lm(an_normalized_amplitude ~ year, data = tmp)
  print(plot(lm1, which = 1, main = st))
  print(plot(lm1, which = 2, main = st))
  print(acf(resid(lm1), main = st))
    cat("-----------------------")
  cat("-----------------------")
}
```


### Water temperature  

```{r, results = FALSE, fig.height = 7, fig.width = 5}
op <- par()
par(mfrow = c(2, 2))

tmp2 <- tmp_annual |> unnest(cols = data)
for(i in seq_along(unique(tmp2$station))){
  st = unique(tmp2$station)[i]
  tmp <- tmp2 |> filter(station == st)
  print(plot(an_normalized_amplitude ~ year, data = tmp, main = paste(st, "time series"), type = "b"))
  lm1 <- lm(an_normalized_amplitude ~ year, data = tmp)
  print(plot(lm1, which = 1, main = st))
  print(plot(lm1, which = 2, main = st))
  print(acf(resid(lm1), main = st))
    print("-----------------------")
  print("-----------------------")
}
```